---
title: "Analysis Pipeline: Dynamic Truncation"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

- Reliable estimates of epidemiological distributions are required for many applications in real-time. Examples include estimates of the incubation period, the time from onset to report, and the time from onset to death. These quantities are typically assumed to follow parametric distributions but may vary over time and due to the characteristics of cases
- The data used  to estimate these distributions can suffer from a range of common biases due to censoring and truncation.
- These are... 
- These are a particular issues for infectious diseases due to the exponential nature of transmission...

- What have other people done? 
  - Examples of good practice
  - Examples of bad practice (maybe?)
  - Options for tools to address this issue


- In this study we explore the impact of these biases on naive estimates of a distribution during a simulated outbreak. We then evaluate a range of approaches to mitigate these biases and compare and contrast there performance. In particular, we focus on the role of right-truncation and dynamic adjustments to explore the relative advantages as well as if they should be used together.
We then apply these approaches to a case study.... and discuss the difference in estimates. We aim to highlight the implications of common biases found when estimating epidemiological distributions and suggest approaches to mitigate them.

# Pipeline

The analysis pipeline for this work can be regenerated by rendering this file,

```{r, eval = FALSE}
rmarkdown::render("_targets.Rmd")
```

The pipeline can then be run using,

```{r, eval = FALSE}
tar_make()
```

The complete pipeline can be visualised using,

```{r, eval = FALSE}
tar_visnetwork()
```

Alternatively the pipeline can be explored interactively using this notebook or updated programmatically using the scripts in `bin`. We also provide an archived version of our `targets` workflow if only wanting to reproduce sections of our analysis. This can be downloaded using the following,

```{r, eval = FALSE}
source(here::here("R", "targets-archive.R"))
get_targets_archive()
```

# Setup

Set up the workflow pipeline and options. We first load the `targets` package and remove the potentially outdated workflow.

```{r}
library(targets)
library(stantargets)
library(tarchetypes)
library(data.table)
library(ggplot2)
library(purrr, quietly = TRUE)
library(here)
library(future)
library(future.callr)
tar_unscript()
```

We now define shared global options across our workflow and load R functions from the `R` folder.

```{targets globals, tar_globals = TRUE}
library(targets)
library(tarchetypes)
library(cmdstanr)
library(data.table)
library(ggplot2)
library(purrr, quietly = TRUE)
library(here)
library(future)
library(future.callr)
plan(callr)
functions <- list.files(here("R"), full.names = TRUE)
walk(functions, source)
rm("functions")
set_cmdstan_path()  

tar_option_set(
  packages = c("data.table", "ggplot2", "purrr", "cmdstanr", "brms", "here"),
  deployment = "worker",
  memory = "transient",
  workspace_on_error = TRUE,
  error = "continue",
  garbage_collection = TRUE
)
```

# Methods

## Models

We explore a range of models for estimating the log normal distribution. Starting with the naive continuous model and then gradually adding complexity to adjust for censoring to dates and right truncation. 

```{targets models, tar_simple = TRUE}
list(
  "Naive" = naive_delay,
  "Censoring adjusted" = censoring_adjusted_delay,
  "Filtered" = function(data, ...) {
    data <- data |>
      data.table::copy() |>
      DT(stime_daily <= (obs_at - 15))
    naive_delay(data = data, ...)
  },
  "Filtered and censoring adjusted" = function(data, ...) {
    data <- data |>
      data.table::copy() |>
      DT(stime_daily <= (obs_at - 15))
    censoring_adjusted_delay(data = data, ...)
  },
  "Truncation adjusted" = truncation_adjusted_delay,
  "Truncation and censoring adjusted" = truncation_censoring_adjusted_delay,
  "Latent variable truncation and censoring adjusted" =
    latent_truncation_censoring_adjusted_delay
)
```

- File path for each saved model.

```{targets model_paths}
tar_target(
  model_paths,
  paste0(
    "data/models/", gsub(" ", "_", tolower(names(models))), ".stan"
  ) |>
    fs::file_create(),
  format = "file",
  pattern = map(models)
)
```

- Dummy data required for model creation.

```{targets dummy_obs, tar_simple = TRUE}
data.table(
  ptime = 1, stime = 2, delay_daily = 1, delay_lwr = 1, delay_upr = 2, 
  ptime_lwr = 1, ptime_upr = 2, stime_lwr = 1, stime_upr = 2, obs_at = 100,
  censored = "interval", censored_obs_time = 10, ptime_daily = 1,
  stime_daily = 1
)
```

- Generate stan code for each model and compile.

```{targets model_stan_code}
tar_target(
  model_stan_code, 
  do.call(
    models[[1]],
    list(data = dummy_obs, fn = brms::make_stancode, save_model = model_paths)
  ),
  pattern = map(models, model_paths)
)
```

### Simulation

### Setup

- We assume 3 distribution scenarios: short, medium, and long.

```{targets distributions}
tar_group_by(
  distributions,
    data.table(
    scenario = c("short", "medium", "long"),
    meanlog = c(1.2, 1.6, 2),
    sdlog = c(0.4, 0.6, 0.8)
  ) |>
    DT(, mean := exp(meanlog + sdlog^2/2)) |>
    DT(, sd := exp(meanlog + (1/2)*sdlog^2)*sqrt(exp(sdlog^2) - 1)),
  scenario
)
```

- We simulate an outbreak that starts with an initially high and stable growth rate that then declines linearly until reaching a stable decay rate. **I am now wondering if we want to investigate more than one scenario.** 

```{targets growth_rate, tar_simple = TRUE}
data.table(
  time = 0:59,
  r = c(rep(0.2, 20), 0.2 - 0.02 * 1:20, rep(-0.2, 20))
)
```

- We initialise the outbreak to have 20 cases.

```{targets init_cases, tar_simple = TRUE}
20
```

- Simulate the outbreak. This is temporary and we should fill with a stochastic SIR (? or other).

```{targets simulated_cases, tar_simple = TRUE}
simulate_uniform_cases(sample_size = 1000, t = 60)
```

- Simulate observations of primary and secondary events as linelist for each distribution scenario.

```{targets simulated_secondary}
tar_target(
  simulated_secondary, 
  simulated_cases |>
    simulate_secondary(
      meanlog = distributions[, "meanlog"][[1]],
      sdlog = distributions[, "sdlog"][[1]]
    ) |>
    DT(, distribution := distributions[, "scenario"][[1]]),
  pattern = map(distributions)
)
```

- Simulate the observation process

```{targets simulated_observations, tar_simple = TRUE}
simulated_secondary |>
  observe_process()
```


### Estimate distributions

- Estimate all models at chosen points across the outbreak (suggestion: "early outbreak" (15 days), "near peak" (30 days), "past peak" (45  days), "late outbreak" (60 days))

```{targets estimation_times}
tar_group_by(
  estimation_times,
  data.table(
    scenario = c("early outbreak", "near peak", "past peak", "late outbreak"),
    time = c(15, 30, 45, 60)
  ),
  scenario
)
```

- Truncate the available simulate observations based on the estimation time for each scenario.

```{targets truncated_obs} 
tar_target(
  truncated_obs,
  simulated_observations |>
    filter_obs_by_obs_time(obs_time = estimation_times[, "time"][[1]]) |>
    DT(, estimation_time := estimation_times[, "scenario"][[1]]),
  pattern = map(estimation_times)
)
```

```{targets group_truncated_obs}
tar_group_by(
  group_truncated_obs,
  truncated_obs,
  estimation_time, distribution
)
```

- Estimate across sample size ranges (N = 10, 100, 1000). `N = 1000` is the main case.

```{targets sample_sizes, tar_simple = TRUE}
c(10, 100, 1000)
```

- Sample observations

```{targets sampled_observations}
tar_target(
  sampled_observations,
  group_truncated_obs |>
    as.data.table() |>
    DT(sample(1:.N, sample_sizes, replace = FALSE)) |>
    DT(, sample_size := as.factor(sample_sizes)),
  pattern = cross(sample_sizes, group_truncated_obs)
)
```

```{targets group_sampled_observations}
tar_group_by(
  group_sampled_observations,
  sampled_observations,
  estimation_time, distribution, sample_size
)
```

- Create data for each group and model and then fit models to this data.

- Summarise posteriors for each model and sampled data.

- Sample model runtimes for each sampled data.

- Plot distribution summary parameters (log mean and sd) vs true values (y facet) by outbreak point (x facet).

### Post-process for dynamic bias

- Post process all models using dynamic correction and known growth rate
- Same structure as simple distribution estimates

### Summarise runtimes


- In addition to estimate quality we care about tractibility for the SI plot the runtimes of various methods.

## Case study

### Data

- Need some outbreak linelist data. The UKHSA paper may contain this.
  - Seb suggest some Ebola linelist data that may be public domain.

### Estimate distributions

- Estimate distributions for similar points as in the simulation setting.

### Estimate growth rate

- Estimate the growth rate retrospectively and comment on this. 

### Post-process for dynamic bias

- Apply growth rate correction to all estimates 

### Summarise runtimes

- Summarise


## Results



## Discussion

### Summary

### Strengths and weaknesses

### Literature

### Further work

- Longer or more complex censoring
- Joint estimation of the transmission process and distributions
- Time-varying distributions

## Conclusions

## Ackowledgements

# References
### For discussion ?
