---
title: "Using `epidist` to estimate..."
description: "A more detailed guide to using the `epidist` R package"
output: 
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
    number_sections: true
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting in depth with epidist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# exclude compile warnings from cmdstanr
knitr::opts_chunk$set(
  fig.path = "figures/epidist-",
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

```{r load-requirements}
library(epidist)
library(data.table)
library(purrr)
library(ggplot2)
```

# Data preparation

* Load the Ebola data, explain the columns and some of the context
* Clean it up into the right format
* Visualise the delay data. Perhaps any other relevant features -- we don't have geocoordinates right

# Model fitting

* Fit the lognormal model
* Fit the gamma model
* Other things here? Convergence checking?

# Post-processing

* Extract posterior predictions (of something) from both models -- posterior predictive checking?
* Compare the models. Give suggestion of which is a better fit
* Translate fitted distribution to probability mass function

## Bibliography {-}