---
title: "Using `epidist` to estimate stratified delays between symptom onset and positive test for the 2014-2016 Ebola outbreak in Sierra Leone"
description: "A more detailed guide to using the `epidist` R package"
output: 
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
    number_sections: true
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting in depth with epidist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
# exclude compile warnings from cmdstanr
knitr::opts_chunk$set(
  fig.path = "figures/epidist-",
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

The 2014-2016 outbreak of Ebola in West Africa began in Guinea, and quickly spread to neighbouring Liberia and Sierra Leone.
It was the largest outbreak in the disease's history, and resulted in over 28,600 infections and 11,325 deaths [@who_ebola_2014_2016].

In this vignette, we use the `epidist` package to analyze line list data from the outbreak in Sierra Leone.
This data was collated by @fang2016transmission, and is available in the `epidist` package.
In analyzing this data, we demonstrate the following features of `epidist`:

1. Fitting district-sex stratified delay distribution estimates.
2. Fitting models with lognormal and gamma delay distributions.
3. Selecting between fitted models.
4. Post-processing and plotting functionality.

For users new to `epidist`, before reading this article, we recommend beginning with the "[Getting started vignette](http://epidist.epinowcast.org/articles/epidist.html)".

The packages used in this article are:

```{r load-requirements}
library(epidist)
library(data.table)
library(purrr)
library(ggplot2)
library(sf)
library(tidybayes)
library(modelr)
```

# Data preparation

We begin by loading the Ebola line list data.
For more details, see `?sierra_leone_ebola_data`.

```{r}
data("sierra_leone_ebola_data")
```

The data has `r nrow(sierra_leone_ebola_data)` rows, each corresponding to a unique case report `id`.
The columns of the data are the individuals name (retracted, and hence can be safely removed), age, sex, the dates of symptom onset and positive sample, and their district and chiefdom.

```{r}
sierra_leone_ebola_data <- sierra_leone_ebola_data |>
  dplyr::select(-name) |>
  dplyr::mutate(case = as.integer(id), .keep = "unused")

head(sierra_leone_ebola_data)
```

Figure \@ref(fig:ebola-outbreak) shows the data.

(ref:ebola-outbreak) Primary and secondary event times for a fraction of the cases.

```{r ebola-outbreak, fig.cap="(ref:ebola-outbreak)", fig.height=9}
# Might want to move to using data.table code rather than dplyr here/everywhere
sierra_leone_ebola_data |>
  dplyr::filter(case %% 10 == 0) |>
  ggplot() +
    geom_segment(
      aes(
        x = date_of_symptom_onset, xend = date_of_sample_tested,
        y = case, yend = case
      ),
      col = "grey"
    ) +
    geom_point(aes(x = date_of_symptom_onset, y = case), col = "#56B4E9") +
    geom_point(aes(x = date_of_sample_tested, y = case), col = "#009E73") +
    facet_wrap(district ~ ., ncol = 2) +
    labs(x = "", y = "Case ID") +
    theme_minimal()
```

<!-- Question about whether to include shapefiles as data object in package. -->

We can use a map to help visualise the outbreak across space.
To create a map, we first join the Ebola data to shapefiles (obtained from [GADM](https://gadm.org/)) for the districts of Sierra Leone.

```{r message = FALSE}
sf <- sf::st_read("../inst/gadm41_SLE_shp/gadm41_SLE_2.shp")

sierra_leone_ebola_data_sf <- dplyr::select(sf, district = NAME_2, geometry) |>
  dplyr::left_join(
    sierra_leone_ebola_data |>
      dplyr::group_by(district) |>
      dplyr::summarise(cases = dplyr::n())
  )
```

Figure \@ref(fig:ebola-cloropleth) shows that the majority of cases we concentrated in the Western Urban district.

(ref:ebola-cloropleth) Figure caption.

```{r ebola-cloropleth, fig.cap="(ref:ebola-cloropleth)"}
ggplot(sierra_leone_ebola_data_sf, aes(fill = cases, geometry = geometry)) +
  geom_sf() +
  scale_fill_viridis_c(na.value = "lightgrey") +
  theme_minimal() +
  labs(fill = "Cases")
```

# Fitting district-sex stratified delay distributions

## Data preparation

To prepare the data for use in `epidist`, we first rename the date columns as `ptime` (time of primary event) and `stime` (time of secondary event), as well as transforming to a `data.table`:

```{r}
sierra_leone_ebola_data <- sierra_leone_ebola_data |>
  dplyr::mutate(
    date_of_symptom_onset = lubridate::ymd(date_of_symptom_onset),
    date_of_sample_tested = lubridate::ymd(date_of_sample_tested),
    ptime = as.numeric(date_of_symptom_onset - min(date_of_symptom_onset)),
    stime = as.numeric(date_of_sample_tested - min(date_of_symptom_onset))
  ) |>
  dplyr::select(case, ptime, stime, age, sex, district) |>
  data.table::as.data.table()
```

Next, we use `observe_process()` to do add interval censoring columns:

```{r}
obs_cens <- epidist::observe_process(sierra_leone_ebola_data)

all(obs_cens$ptime == obs_cens$ptime_daily)
all(obs_cens$stime == obs_cens$stime_daily)

max(obs_cens$ptime)
max(obs_cens$stime)
```

The columns added are:

* `ptime_daily`: As the `ptime` was already daily then this is superfluous
* `ptime_lwr`: The lower bound on the primary observation time i.e. the floor of `ptime_daily`
* `ptime_upr`: The upper bound on the primary observation time i.e. `ptime_lwr + 1`
* `stime_daily`: As the `ptime` was already daily then this is superfluous
* `stime_lwr`: The lower bound on the secondary observation time i.e. the floor of `stime_daily`
* `stime_upr`: The upper bound on the secondary observation time i.e. `stime_lwr + 1`
* `delay_daily`: Is given by `stime_daily - ptime_daily`
* `delay_lwr`: Is given by `stime_lwr - ptime_upr` i.e. the minimum possible delay
* `delay_upr`: Is given by `stime_upr - ptime_lwr` i.e. the maximum possible delay
* `obs_at`: The maximum value of `stime`

## Model fitting

Now we fit lognormal delay distribution with sex effect and district effect.
First fit model with no district-sex stratification.
Sub-sample the data for speed.
Use ADVI for speed.

```{r}
set.seed(1)

N <- nrow(obs_cens)
obs_cens <- obs_cens[complete.cases(obs_cens)]
N_complete <- nrow(obs_cens)
N_complete / N
obs_cens <- obs_cens[sample(seq_len(.N), N_complete / 5, replace = FALSE)]
obs_prep <- as_latent_individual(obs_cens)

fit <- epidist::epidist(
  data = obs_prep,
  formula = brms::bf(mu ~ 1, sigma ~ 1),
  family = brms::lognormal(),
  algorithm = "laplace",
  refresh = 0
)

summary(fit)
```

Fit a model including sex effect:

```{r}
fit_sex <- epidist::epidist(
  data = obs_prep,
  formula = brms::bf(mu ~ 1 + sex, sigma ~ 1 + sex),
  family = brms::lognormal(),
  algorithm = "laplace",
  refresh = 0
)

summary(fit_sex)
```

Fit a model including sex effect and district effect:

```{r}
fit_sex_district <- epidist::epidist(
  data = obs_prep,
  formula = brms::bf(mu ~ 1 + sex + district, sigma ~ 1 + sex + district),
  family = brms::lognormal(),
  algorithm = "laplace",
  refresh = 0
)

summary(fit_sex_district)
```

## Posterior expectations

Posterior expectation of the delay distribution by covariate under no censoring or truncation:

```{r}
epred_draws <- obs_prep |>
  as.data.frame() |>
  modelr::data_grid(NA) |>
  dplyr::mutate(obs_t = NA, pwindow_upr = NA, swindow_upr = NA) |>
  add_epred_draws(fit, dpar = TRUE)

epred_draws |>
  ggplot(aes(x = .epred)) +
    tidybayes::stat_slab() +
    labs(x = "Posterior expectation of the delay", y = "") +
    theme_minimal()

# We will probably want to add functionality for users to mutate these things
# to the newdata themselves
epred_draws_sex <- obs_prep |>
  as.data.frame() |>
  modelr::data_grid(sex) |>
  dplyr::mutate(obs_t = NA, pwindow_upr = NA, swindow_upr = NA) |>
  tidybayes::add_epred_draws(fit_sex, dpar = TRUE)

epred_draws_sex |>
  dplyr::mutate(
    sex = dplyr::case_when(
      sex == "F" ~ "Female",
      sex == "M" ~ "Male"
    )
  ) |>
  ggplot(aes(x = .epred, y = sex)) +
    tidybayes::stat_slab() +
    labs(x = "Posterior expectation of the delay", y = "") +
    theme_minimal()

epred_draws_sex_district <- obs_prep |>
  as.data.frame() |>
  modelr::data_grid(sex, district) |>
  dplyr::mutate(obs_t = NA, pwindow_upr = NA, swindow_upr = NA) |>
  add_epred_draws(fit_sex_district, dpar = TRUE)

epred_draws_sex_district |>
  dplyr::mutate(
    sex = dplyr::case_when(
      sex == "F" ~ "Female",
      sex == "M" ~ "Male"
    )
  ) |>
  ggplot(aes(x = .epred, y = district)) +
    tidybayes::stat_slab() +
    facet_grid(. ~ sex) +
    labs(x = "Posterior expectation of the delay", y = "") +
    theme_minimal()

# Some strata that don't have much data? Some very bad priors?
summary(epred_draws_sex_district$.epred)

obs_prep$district |>
  table()

epred_draws_sex_district |>
  dplyr::filter(district %in% c("Port Loko", "Western Urban")) |>
  dplyr::mutate(
    sex = dplyr::case_when(
      sex == "F" ~ "Female",
      sex == "M" ~ "Male"
    )
  ) |>
  ggplot(aes(x = .epred, y = district)) +
    tidybayes::stat_slab() +
    facet_grid(. ~ sex) +
    labs(x = "Posterior expectation of the delay", y = "") +
    theme_minimal()

epred_draws_sex_district |>
  dplyr::group_by(district) |>
  dplyr::summarise(q90 = quantile(.epred, 0.9)) |>
  ggplot(aes(x = district, y = q90)) +
    geom_point()

obs_prep$district |>
  table()
```

## Posterior predictions

Posterior prediction of the delay distribution under censoring and truncation.
Because we are copying the data.
Then plot the data on top of the posterior predictive distribution.
Does it look reasonable?

## Delay probability mass functions

Do a posterior predictive distribution under no truncation but daily censoring.
Then plot those as discrete PMF.
Compare to data somehow?

What about the "continuous" no truncation no censoring distribution?

## Model comparison

* Is it worth including sex as a covariate? 
* Is it worth including district as a covariate? What about spatially structured?

# Using a gamma delay distribution

## Fitting 

* Just fit the best model from above here?
* Going to do the sex stratified one for now

```{r}
fit_sex_gamma <- epidist::epidist(
  data = obs_prep,
  formula = brms::bf(mu ~ 1 + sex, shape ~ 1 + sex),
  family = stats::Gamma(),
  algorithm = "laplace",
  refresh = 0
)

summary(fit_sex_gamma)
```

# Model selection

* Should we use the lognormal or gamma delay distribution?

# Conclusion

## Bibliography {-}