---
title: "Getting started with epidist"
description: "A quick start guide to using the epidist R package"
output: 
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
    number_sections: true
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with epidist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
# exclude compile warnings from cmdstanr
knitr::opts_chunk$set(
  fig.path = "figures/epidist-",
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

```{r load-requirements}
library(epidist)
library(data.table)
library(purrr)
library(ggplot2)
```

Many quantities in epidemiology can be thought of as the time between two events, or "delays".
Important examples include:

* the incubation period (time between infection and symptom onset),
* serial interval (time between symptom onset of infectee and symptom onset of infected), and 
* generation interval (time between infection of infectee and infection of infected).

Unfortunately, estimating delays accurately from observational data is usually difficult.
In our experience, the two main issues are:

1. interval censoring, and
2. right truncation.

Don't worry if you've not come across these terms before.
In Section \@ref(data), we will explain what they mean by simulating data like that we might observe during an ongoing infectious disease outbreak.
Next, in Section \@ref(fit), we show how `epidist` can be used to estimate delays using a statistical model which properly accounts for these two issues.
Finally, in Section \@ref(compare), we demonstrate that the fitted delay distribution accurately recovers the underlying truth.

If you would like more technical details, the `epidist` package implements models following best practices as described in @park2024estimating and @charniga2024best.

# Example data {#data}

Data should be formatted as a [`data.table`]() with the following columns for use within the `epidist` package:

* `case`: 
* `ptime`:
* `stime`:

Here we simulate data in this format, and in doing so explain the two main issues with observational delay data.
<!-- Please see as of yet unwritten vignette for information about how to transform your data of other formats to the right format. -->

First, we use the [Gillepsie algorithm](https://en.wikipedia.org/wiki/Gillespie_algorithm) to generate infectious disease outbreak data (Figure \@ref(fig:outbreak)) from a stochastic compartmental model:

```{r}
outbreak <- simulate_gillespie(seed = 101)
```

(ref:outbreak) Early on in the epidemic, there is a high rate of growth in new cases. As more people are infected, the rate of growth slows. (Only every 50th case is shown to avoid over-plotting.)

```{r outbreak, fig.cap="(ref:outbreak)"}
outbreak[case %% 50 == 0, ] |>
  ggplot(aes(x = ptime, y = case)) +
  geom_point(col = "#56B4E9") +
  labs(x = "Primary event time (day)", y = "Case number") +
  theme_minimal()
```

`outbreak` is a [`data.table`]() with the columns `case` and `ptime`.

Now, to generate secondary events, we will use a lognormal distribution (Figure \@ref(fig:lognormal)) for the delay between primary and secondary events:

```{r}
secondary_dist <- data.table(meanlog = 1.8, sdlog = 0.5) |>
  add_natural_scale_mean_sd()
```

(ref:lognormal) The lognormal distribution is skewed to the right. Long delay times still have some probability.

```{r lognormal, fig.cap="(ref:lognormal)"}
ggplot(data.frame(x = c(0, 8)), aes(x = x)) +
  geom_function(
    fun = dlnorm,
    meanlog = secondary_dist[["meanlog"]],
    sdlog = secondary_dist[["sdlog"]]
  ) +
  theme_minimal() +
  labs(
    x = "Delay between primary and secondary event (days)",
    y = "Probability density"
  )
```

```{r}
obs <- outbreak |>
  simulate_secondary(
    meanlog = secondary_dist[["meanlog"]],
    sdlog = secondary_dist[["sdlog"]]
  )
```

(ref:delay) Secondary events (in green) occur with a delay drawn from the lognormal distribution (Figure \@ref(fig:lognormal)). As with Figure \@ref(fig:outbreak), to make this figure easier to read, only every 50th case is shown.

```{r delay, fig.cap="(ref:delay)"}
obs[case %% 50 == 0, ] |>
  ggplot(aes(y = case)) +
  geom_segment(
    aes(x = ptime, xend = stime, y = case, yend = case),
    col = "grey"
  ) +
  geom_point(aes(x = ptime), col = "#56B4E9") +
  geom_point(aes(x = stime), col = "#009E73") +
  labs(x = "Event time (day)", y = "Case number") +
  theme_minimal()
```

`obs` is now a [`data.table`]() object with further columns for `delay` and `stime`.
The secondary event time is simply the primary event time plus the delay:

```{r}
all(obs$ptime + obs$delay == obs$stime)
```

If we were to receive the data `obs` as above then estimating the delay distribution would be easy, and the `epidist` package wouldn't need to exist.
However, in reality, during an outbreak we almost never receive the data as above.

First, the times of primary and secondary events will usually be censored.
This means that rather than exact event times, we observe event times within an interval.
Here we suppose that the interval is daily, meaning that only the date of the primary or secondary event, not the exact event time, is reported (Figure \@ref(fig:cens)):

```{r}
# observe_process() should be renamed and include a "daily" i.e. 1 argument
obs_cens <- obs |> observe_process()
```

(ref:cens) Interval censoring of the primary and secondary event times obscures the delay times.

```{r cens, fig.cap="(ref:cens)"}
ggplot(obs_cens, aes(x = delay, y = delay_daily)) +
  geom_point(col = "#E69F00") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "grey") +
  theme_minimal() +
  coord_fixed() +
  labs(x = "Exact delay time (days)", y = "Censored delay time (days)")
```

Next, during an outbreak we will usually be estimating delays in real time.
The result is that only those cases with a secondary event occurring before some time will be observed.
This is called (right) truncation, and biases the observation process towards shorter delays:
<!-- I think possible to do a better job with showing what right truncation is with a figure here! -->

```{r}
obs_time <- 25
# filter_obs_by_obs_time() should be renamed to refer to stime
obs_cens_trunc <- filter_obs_by_obs_time(obs_cens, obs_time = obs_time)
```

Finally, in reality, it's not possible to observe every case.
We suppose that a sample of individuals of size `sample_size` are observed:

```{r}
sample_size <- 200
```

This sample size corresponds to `r 100 * round(sample_size / nrow(obs_cens_trunc), 3)`% of the data.

```{r}
obs_cens_trunc_samp <-
  obs_cens_trunc[sample(seq_len(.N), sample_size, replace = FALSE)]
```

With our censored, truncated, and sampled data, we are now ready to try to recover the underlying delay distribution using `epidist`.

# Fit the model {#fit}

If we had access to `obs`, then it would be simple to estimate the delay distribution.
However, with only access to `obs_cens_trunc_samp`, we must use a statistical model to avoid severely biased estimates!

The main model function in `epidist` is `latent_truncation_censoring_adjusted_delay`.
<!-- In a future vignette, we will explain in more detail the structure of the model. -->
The parameters of this model are inferred using Bayesian inference.
In particular, we use the the No-U-Turn Sampler (NUTS) Markov chain Monte Carlo (MCMC) algorithm via the `brms` R package.

```{r}
fit <- latent_truncation_censoring_adjusted_delay(
  data = obs_cens_trunc_samp, cores = 4, refresh = 0
)

summary(fit)
```

# Compare estimates {#compare}

Compare modelled estimate to underlying truth here to convince user that it works.
Perhaps as an exercise, we could show the "really simple" estimate being wrong here otherwise it won't be so impressive that the model is right.
This section should also show the user how to get objects that they might want out of the fitted object.

```{r}
draws <- extract_lognormal_draws(fit)
draws_long <- draws_to_long(draws)
summarise_draws(draws_long, sf = 2)
```

## Bibliography {-}
