---
title: "Extending epidist"
description: "Building your own model type, and the packages that already do"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Extending epidist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Why extend `epidist`?

`epidist` handles the parts of delay estimation that are tedious and easy to get wrong.
Censoring, truncation, the `brms` formula interface, priors, and the Stan code that ties them together.

If you are building a model that shares those problems but has a different likelihood, you can reuse all of it rather than rewriting it.
Your package supplies the likelihood and the Stan code specific to your model.
`epidist` supplies everything around it, and your users get the `brms` formula interface for free.

Two packages already do this, in two different ways.

```{r packages-table, eval = TRUE, echo = FALSE}
knitr::kable(
  data.frame(
    Package = c("cfrnow", "tbl.now"),
    Extends = c("Model type", "Data source"),
    What = c(
      paste(
        "Real-time case fatality ratio via a Bayesian mixture-cure",
        "survival model. The CFR and the onset-to-death delay both take",
        "brms formulas."
      ),
      paste(
        "A tidy nowcast tibble class, with methods so it converts into",
        "epidist's own data classes."
      )
    ),
    stringsAsFactors = FALSE
  ),
  col.names = c("Package", "Extends", "What it does")
)
```

- [`cfrnow`](https://github.com/epiforecasts/cfrnow) registers a new **model type**.
- [`tbl.now`](https://github.com/RodrigoZepeda/tbl.now) adds a new **data source**, by writing `as_epidist_linelist_data()` and `as_epidist_aggregate_data()` methods for its own class.

# Adding a data source

This is the smaller job.
Write methods for `as_epidist_linelist_data()` or `as_epidist_aggregate_data()` that turn your class into one `epidist` understands.
Nothing else is needed, because from that point your data is an ordinary `epidist` object.

# Adding a model type

A model type is an S3 class plus methods for six generics.
`cfrnow` implements exactly these and nothing else, so this is the whole extension interface:

| Generic | What you supply |
|---|---|
| `assert_epidist()` | Validation. What columns your data must have, and what must be true of them |
| `epidist_family_model()` | The `brms` custom family, including the likelihood name and its `vars` |
| `epidist_formula_model()` | How the user's formula is rewritten, usually adding `vreal()` terms |
| `epidist_transform_data_model()` | Any reshaping needed before fitting. Return the data unchanged if none |
| `epidist_model_prior()` | Model specific priors, or `NULL` |
| `epidist_stancode()` | The Stan code injected through `brms` `stanvars` |

You also write a constructor that sets your class, by convention `as_epidist_<name>_model()`.

The skeleton looks like this.

```{r skeleton}
as_epidist_my_model <- function(data) {
  class(data) <- c("epidist_my_model", class(data))
  assert_epidist(data)
  return(data)
}

#' @importFrom epidist assert_epidist
#' @export
assert_epidist.epidist_my_model <- function(data, ...) {
  checkmate::assert_names(names(data), must.include = c("delay", "n"))
  return(invisible(NULL))
}

#' @importFrom epidist epidist_family_model
#' @export
epidist_family_model.epidist_my_model <- function(data, family, ...) {
  return(brms::custom_family(
    paste0("my_model_", family$family),
    dpars = family$dpars,
    vars = "vreal1",
    loop = FALSE
  ))
}

#' @importFrom epidist epidist_formula_model
#' @export
epidist_formula_model.epidist_my_model <- function(data, formula, ...) {
  return(stats::update(formula, delay | vreal(n) ~ .))
}

#' @importFrom epidist epidist_stancode
#' @export
epidist_stancode.epidist_my_model <- function(data, family, formula, ...) {
  return(brms::stanvar(block = "functions", scode = "// your lpdf here"))
}
```

Register the S3 methods in your `NAMESPACE` and import the generics from `epidist`, as `cfrnow` does:

```
importFrom(epidist,assert_epidist)
importFrom(epidist,epidist_family_model)
S3method(assert_epidist,epidist_my_model)
S3method(epidist_family_model,epidist_my_model)
```

Users then call `epidist()` exactly as they would for a built-in model.

```{r use}
data |>
  as_epidist_my_model() |>
  epidist(formula = mu ~ 1)
```

# A worked example

`cfrnow` is the clearest one to read.
Its [`R/cure_model.R`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R) is about two hundred lines and implements the six generics above for a mixture-cure survival model.
It is a good template because the model is genuinely different from the built-in ones, yet almost all the code is about its own likelihood rather than about `epidist`.

# Getting help

If you are building an extension and something in this interface is awkward, please open an issue.
The generics above are the supported surface, and we would rather change them than have you work around them.
