---
title: "Extending epidist"
description: "Building your own model type"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Extending epidist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Why extend `epidist`?

`epidist` handles the parts of delay estimation that are tedious and easy to get wrong.
Censoring, truncation, the `brms` formula interface, priors, and the Stan code that ties them together.

If you are building a model that shares those problems but has a different likelihood, you might want to do so by extending `epidist`.

Two packages already do this, in two different ways.

- [`cfrnow`](https://github.com/epiforecasts/cfrnow) registers a new **model type**.
- [`tbl.now`](https://github.com/RodrigoZepeda/tbl.now) adds a new **data source**, by writing `as_epidist_linelist_data()` and `as_epidist_aggregate_data()` methods for its own class.

# Adding a data source

Write a method that turns your class into one `epidist` understands.

```{r data-source}
#' @importFrom epidist as_epidist_linelist_data
#' @export
as_epidist_linelist_data.my_class <- function(data, ...) {
  return(as_epidist_linelist_data(
    as.data.frame(data),
    pdate_lwr = "onset_date",
    sdate_lwr = "report_date"
  ))
}
```

```{r data-source-use}
my_data |>
  as_epidist_linelist_data() |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1)
```

# Adding a model type

A model type is an S3 class plus methods for six generics.

Three you have to supply.

| Generic | What you supply | If you skip it |
|---|---|---|
| `assert_epidist()` | What columns your data must have and what must be true of them | Errors |
| `epidist_family_model()` | The `brms` custom family, including the likelihood name and its `vars` | Returns the family unchanged |
| `epidist_stancode()` | The Stan code injected through `brms` `stanvars` | `NULL`, so no likelihood |

Three are optional.

| Generic | What you supply | If you skip it |
|---|---|---|
| `epidist_formula_model()` | How the user's formula is rewritten, usually adding `vreal()` terms | Returns the formula unchanged |
| `epidist_transform_data_model()` | Reshaping before fitting | Returns the data unchanged |
| `epidist_model_prior()` | Model specific priors | `NULL` |

You also write a constructor that sets your class, by convention `as_epidist_<name>_model()`.

The skeleton looks like this.

```{r skeleton}
as_epidist_my_model <- function(data) {
  class(data) <- c("epidist_my_model", class(data))
  assert_epidist(data)
  return(data)
}

#' @importFrom epidist assert_epidist
#' @export
assert_epidist.epidist_my_model <- function(data, ...) {
  checkmate::assert_names(names(data), must.include = c("delay", "n"))
  return(invisible(NULL))
}

#' @importFrom epidist epidist_family_model
#' @export
epidist_family_model.epidist_my_model <- function(data, family, ...) {
  return(brms::custom_family(
    paste0("my_model_", family$family),
    dpars = family$dpars,
    vars = "vreal1",
    loop = FALSE
  ))
}

#' @importFrom epidist epidist_formula_model
#' @export
epidist_formula_model.epidist_my_model <- function(data, formula, ...) {
  return(stats::update(formula, delay | vreal(n) ~ .))
}

#' @importFrom epidist epidist_stancode
#' @export
epidist_stancode.epidist_my_model <- function(data, family, formula, ...) {
  return(brms::stanvar(block = "functions", scode = "// your lpdf here"))
}
```

Register the S3 methods in your `NAMESPACE` and import the generics from `epidist`, as `cfrnow` does:

```
importFrom(epidist,assert_epidist)
importFrom(epidist,epidist_family_model)
importFrom(epidist,epidist_formula_model)
importFrom(epidist,epidist_stancode)
S3method(assert_epidist,epidist_my_model)
S3method(epidist_family_model,epidist_my_model)
S3method(epidist_formula_model,epidist_my_model)
S3method(epidist_stancode,epidist_my_model)
```

Users then call `epidist()` exactly as they would for a built-in model.

```{r use}
data |>
  as_epidist_my_model() |>
  epidist(formula = mu ~ 1)
```

# Exploring how cfrnow extends epidist

[`cfrnow`](https://github.com/epiforecasts/cfrnow) fits a mixture-cure survival model to estimate a real-time case fatality ratio.
Every method it needs is in [`R/cure_model.R`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R).

| Method | |
|---|---|
| [`as_epidist_cure_model()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L49) | Sets the class |
| [`assert_epidist()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L62) | Checks the columns its likelihood needs |
| [`epidist_family_model()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L93) | Declares the custom family and its `vars` |
| [`epidist_formula_model()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L116) | Adds the `vreal()` terms |
| [`epidist_transform_data_model()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L122) | Reshapes before fitting |
| [`epidist_model_prior()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L129) | Returns `NULL` |
| [`epidist_stancode()`](https://github.com/epiforecasts/cfrnow/blob/main/R/cure_model.R#L176) | Injects the mixture-cure likelihood |

`cfrnow` also exports its own frontend.
Its users call `prepare_cfr_data()`, `fit_cfr()` and `pp_check_cfr()` rather than `epidist()`.

# A worked example inside the package: the meta model

The meta model (`vignette("meta")`) is the same exercise carried out inside `epidist`, so its source is a second worked example alongside `cfrnow`.
It fits published summary estimates, which need a different likelihood from individual level delays, and it needs a new data class to hold them.
It is experimental, and its interface may still change.

| Piece | Where | What it does |
|---|---|---|
| `as_epidist_estimates_data()` | `R/estimates_data.R` | A new data class for published summaries, with an `assert_epidist()` method |
| `as_epidist_meta_model()` | `R/meta_model.R` | Sets the model class, from individual data, summaries, or both |
| `epidist_family_model()` | `R/meta_model.R` | Declares the custom family and its `vars` |
| `epidist_formula_model()` | `R/meta_model.R` | Adds the `vint()` and `vreal()` terms |
| `epidist_transform_data_model()` | `R/meta_model.R` | Aggregates individual rows and leaves summary rows alone |
| `epidist_stancode()` | `R/meta_model.R` | Injects the likelihood and the data arrays it indexes into |
| `epidist_gen_meta_log_lik()`, `epidist_gen_meta_predict()` | `R/meta_summaries.R` | Post-processing, so `log_lik()` and `posterior_predict()` work |

Three things it does are worth copying if your own likelihood needs them.

- **Two kinds of row in one model.** Individual level rows reuse the marginal model likelihood and summary rows use their own. A slot on each row says which, and the likelihood branches on it.
- **Data a row cannot carry.** A study may report any number of quantiles, and a row has a fixed number of slots. The reported values are passed to Stan as flat arrays through `brms::stanvar()`, and each row holds a start index and a length into them. Reordering or aggregating rows cannot break that, because the index does not depend on row order.
- **Keeping the R and Stan likelihoods in step.** Post-processing needs an R mirror of the Stan likelihood. `tests/testthat/test-int-meta_model.R` fits a generated quantities program that calls the Stan function directly and checks it against the R version for every kind of row, so the two cannot drift apart.

# Getting help

If you are building an extension and something in this interface is awkward, please open an issue.
