---
title: "Frequently asked questions and tips"
output: 
  bookdown::html_document2:
    toc: true
    fig_caption: yes
    code_folding: show
    number_sections: false
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

Here we provide tips for working with the `epidist` package, and answers to "frequently" asked questions.
If you have a question about using the package, please create an [issue](https://github.com/epinowcast/epidist/issues) and we will endeavour to get back to you soon!

The code block below is provided to facilitate reproduction of this script, if required!

```{r message=FALSE, results='hide', class.source='fold-hide'}
library(epidist)
library(brms)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)
library(tidybayes)

set.seed(1)

meanlog <- 1.8
sdlog <- 0.5
obs_time <- 25
sample_size <- 200

obs_cens_trunc_samp <- simulate_gillespie(seed = 101) |>
  simulate_secondary(
    meanlog = meanlog,
    sdlog = sdlog
  ) |>
  observe_process() |>
  filter_obs_by_obs_time(obs_time = obs_time) |>
  slice_sample(n = sample_size, replace = FALSE)

data <- as_latent_individual(obs_cens_trunc_samp)
fit <- epidist(
  data,
  formula = bf(mu ~ 1, sigma ~ 1),
  seed = 1
)
```

## I would like to work with the samples output

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the [`posterior`](https://mc-stan.org/posterior/) package for working with samples from MCMC or other sampling algorithms.
For example, the function `posterior::as_draws_df()` may be used to obtain a dataframe of MCMC draws for specified parameters.

```{r message = FALSE}
library(posterior)
draws <- as_draws_df(fit, variable = c("Intercept", "Intercept_sigma"))
head(draws)
```

## How can I assess if sampling has converged?

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the [`bayesplot`](http://mc-stan.org/bayesplot/) package for sampling diagnostic plots.
For example, the function `bayesplot::mcmc_trace()` can be used to produce traceplots for specified parameters.

```{r message = FALSE}
library(bayesplot)
mcmc_trace(fit, pars = c("Intercept", "Intercept_sigma"))
```

We also provide a function `epidist_diagnostics()` which can be used to obtain common diagnostics used to assess the quality of a fitted model.

```{r message = FALSE}
epidist_diagnostics(fit)
```

## I'd like to run a simulation study

We recommend use of the [`purrr`](https://purrr.tidyverse.org/) package for running many `epidist` models, for example as a part of a simulation study.
We particularly highlight two functions which might be useful:

1. `purrr::map()` (and other similar functions) for iterating over a list of inputs.
2. `purrr::safely()` which ensures that the function called "always succeeds". In other words, if there is an error it will be captured and output, rather than ending computation (and potentially disrupting a call to `purrr::map()`).

For an example use of these functions, have a look at the [`epidist-paper`](https://github.com/parksw3/epidist-paper) repository containing the code for @park2024estimating.
(Note that in that codebase, we use `map` as a part of a [`targets`](https://books.ropensci.org/targets/) pipeline.)

## How did you choose the default priors for `epidist`?

[`brms`](http://paulbuerkner.com/brms/) provides default priors for all parameters.
However, some of those priors do not make sense in the context of our application.
Instead, we used [prior predictive checking](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html) to set `epidist`-specific default priors which produce epidemiological delay distribution mean and standard deviation parameters in a reasonable range.

For example, for the `brms::lognormal()` latent individual model, we suggest the following prior distributions for the `brms` `mu` and `sigma` intercept parameters:

```{r}
family <- "lognormal"

epidist_family <- epidist_family(data, family)
epidist_formula <- epidist_formula(
  data, family = epidist_family, formula = bf(mu ~ 1, sigma ~ 1)
)

# NULL here means no replacing priors from the user!
epidist_prior <- epidist_prior(
  data = data,
  family = family,
  formula = epidist_formula,
  prior = NULL
)

epidist_prior
```

(Note that the functions `epidist_family()` and `epidist_prior()` are mostly for internal use!)

Here are the distributions on the delay distribution mean and standard deviation parameters that these prior distributions imply:

```{r message=FALSE, results='hide'}
set.seed(1)
fit_ppc <- epidist(
  data = data,
  formula = bf(mu ~ 1, sigma ~ 1),
  family = "lognormal",
  sample_prior = "only",
  seed = 1
)
```

```{r}
pred <- predict_delay_parameters(fit_ppc)

pred |>
  as.data.frame() |>
  pivot_longer(
    cols = c("mu", "sigma", "mean", "sd"),
    names_to = "parameter",
    values_to = "value"
  ) |>
  filter(parameter %in% c("mean", "sd")) |>
  ggplot(aes(x = value, y = after_stat(density))) +
  geom_histogram() +
  facet_wrap(. ~ parameter, scales = "free") +
  labs(x = "", y = "Density") +
  theme_minimal() +
  scale_x_log10(labels = comma)

quantile(pred$mean, c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99))
quantile(pred$sd, c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99))
```

# How can I assess how sensitive the fitted posterior distribution is to the prior distribution used?

We recommend use of the [`priorsense`](https://github.com/n-kall/priorsense) package [@kallioinen2024detecting] to check how sensitive the posterior distribution is to perturbations of the prior distribution and likelihood using power-scaling analysis:

```{r}
library(priorsense)
powerscale_plot_dens(fit, variable = c("Intercept", "Intercept_sigma")) +
  theme_minimal()
```

# What do the parameters in my model output correspond to?

The `epidist` package uses [`brms`](http://paulbuerkner.com/brms/) to fit models.
This means that the model output will include `brms`-style names for parameters.
Here, we provide a table giving the correspondence between the distributional parameter names used in `brms` and those used in standard R functions for some common likelihood families.

| Family        | `brms` parameter | R parameter |
|---------------|------------------|-------------|
| `lognormal()` | `mu`             | `meanlog`   |
| `lognormal()` | `sigma`          | `sdlog`     |
| `Gamma()`     | `mu`             | `shape / scale` |
| `Gamma()`     | `shape`          | `shape`     |

Note that all families in `brms` are parameterised with some measure of centrality `mu` as their first parameter.
This parameter does not necessarily correspond to the mean: hence the provision of a function `add_mean_sd()` within `epidist` to add columns containing the natural scale mean and standard deviation to a `data.frame` of draws.

# How can I generate predictions with my fitted `epidist` model?

It is possible to generate predictions manually by working with [samples from the model output](https://epidist.epinowcast.org/articles/faq.html#i-would-like-to-work-with-the-samples-output).
However this is tricky to do, and so where possible we recommend using the [`tidybayes`](http://mjskay.github.io/tidybayes/) package.
In particular, following functions may be useful:

1. `tidybayes::add_epred_draws()` for predictions of the expected value of a delay.
2. `tidybayes::add_linpred_draws()` for predictions of the delay distributional parameter linear predictors.
3. `tidybayes::add_predicted_draws()` for predictions of the observed delay.

To see these functions demonstrated in a vignette, see ["Advanced features with Ebola data"](https://epidist.epinowcast.org/articles/ebola.html).
As a short example, to generate 4000 predictions (equal to the number of draws) of the delay that would be observed with a double censored observation process (in which the primary and secondary censoring windows are both one) then:

```{r}
draws_pmf <- data.frame(relative_obs_time = 1000, pwindow = 1, swindow = 1) |>
  add_predicted_draws(fit, ndraws = 4000)

ggplot(draws_pmf, aes(x = .prediction)) +
  geom_bar(aes(y = after_stat(count / sum(count)))) +
  labs(x = "Delay", y = "PMF") +
  scale_x_continuous(limits = c(0, 30)) +
  theme_minimal()
```

Importantly, this functionality is only available for `epidist` models using custom `brms` families that have `posterior_predict` and `posterior_epred` methods implemented.
For example, for the `latent_individual` model, currently methods are implemented for the [lognormal](https://github.com/epinowcast/epidist/blob/main/R/latent_lognormal.R) and [gamma](https://github.com/epinowcast/epidist/blob/main/R/latent_gamma.R) families.
If you are using another family, consider [submitting a pull request](https://github.com/epinowcast/epidist/pulls) to implement these methods!
In doing so, you may find it useful to use the [`primarycensored`](https://primarycensored.epinowcast.org/) package.

## Bibliography
