---
title: "Frequently asked questions and tips"
output:
  bookdown::html_document2:
    toc: true
    fig_caption: yes
    code_folding: show
    number_sections: false
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

Here we provide tips for working with the `epidist` package, and answers to frequently asked questions.
If you have a question about using the package, please create an [issue](https://github.com/epinowcast/epidist/issues).
We will endeavour to respond promptly!

The code block below facilitates reproduction of this script.

```{r message=FALSE, results='hide', class.source='fold-hide'}
library(epidist)
library(brms)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)

set.seed(1)

# Define different parameters for each location
meanlog_0 <- 1.6
sdlog_0 <- 0.4
meanlog_1 <- 2.1
sdlog_1 <- 0.6

obs_time <- 25
sample_size <- 200

# Create base data with location assignments
base_data <- simulate_gillespie(seed = 101)
base_data$location <- rbinom(n = nrow(base_data), size = 1, prob = 0.5)

# Create location 0 data
location_0_data <- base_data |>
  filter(location == 0) |>
  simulate_secondary(
    meanlog = meanlog_0,
    sdlog = sdlog_0
  ) |>
  select(case, ptime, delay, stime, location)

# Create location 1 data
location_1_data <- base_data |>
  filter(location == 1) |>
  simulate_secondary(
    meanlog = meanlog_1,
    sdlog = sdlog_1
  ) |>
  select(case, ptime, delay, stime, location)

# Combine datasets
obs_cens_trunc_samp <- bind_rows(location_0_data, location_1_data) |>
  mutate(
    ptime_lwr = floor(.data$ptime),
    ptime_upr = .data$ptime_lwr + 1,
    stime_lwr = floor(.data$stime),
    stime_upr = .data$stime_lwr + 1,
    obs_time = obs_time
  ) |>
  filter(.data$stime_upr <= .data$obs_time) |>
  slice_sample(n = sample_size, replace = FALSE)

linelist_data <- as_epidist_linelist_data(
  obs_cens_trunc_samp$ptime_lwr,
  obs_cens_trunc_samp$ptime_upr,
  obs_cens_trunc_samp$stime_lwr,
  obs_cens_trunc_samp$stime_upr,
  obs_time = obs_cens_trunc_samp$obs_time,
  covariates = data.frame(location = obs_cens_trunc_samp$location)
)
data <- as_epidist_marginal_model(linelist_data)

fit <- epidist(
  data,
  formula = mu ~ 1,
  seed = 1,
  chains = 2,
  cores = 2,
  refresh = ifelse(interactive(), 250, 0),
  iter = 1000,
  backend = "cmdstanr"
)

fit_location <- epidist(
  data,
  formula = bf(mu ~ location, sigma ~ location),
  seed = 1,
  chains = 2,
  cores = 2,
  refresh = ifelse(interactive(), 250, 0),
  iter = 1000,
  backend = "cmdstanr"
)
```

## Working with posterior samples

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the [`posterior`](https://mc-stan.org/posterior/) package for working with samples.

### Getting a `data.frame` of posterior samples

The function `posterior::as_draws_df()` may be used to obtain a dataframe of MCMC draws for specified parameters.

```{r message = FALSE}
library(posterior)
draws <- as_draws_df(fit, variable = c("Intercept", "Intercept_sigma"))
head(draws)
```

### Using random variables (`rvars`) for uncertainty propagation

The `posterior` package also provides the `rvar` class for working with posterior samples as random variables.
This approach allows you to perform mathematical operations directly with posterior distributions while propagating uncertainty.

```{r}
library(posterior)
# Convert model parameters to rvars
rv <- as_draws_rvars(fit_location)

# Look at the intercept in rvar form
rv$b_Intercept

# Calculate the mean delay (on natural scale) for each location
mean_delay_loc0 <- exp(rv$b_Intercept + 0.5 * rv$b_sigma_Intercept^2)
mean_delay_loc1 <- exp(rv$b_Intercept + rv$b_location +
                       0.5 * (rv$b_sigma_Intercept + rv$b_sigma_location)^2)

# Summarize the distributions
summary(mean_delay_loc0)
summary(mean_delay_loc1)

# Calculate the difference between locations
delay_diff <- mean_delay_loc1 - mean_delay_loc0
delay_diff
```

For more details, see the [`posterior` package documentation](https://mc-stan.org/posterior/articles/rvar.html).

## How can I assess whether sampling has converged?

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the [`bayesplot`](http://mc-stan.org/bayesplot/) package for sampling diagnostic plots.
For example, the function `bayesplot::mcmc_trace()` can be used to produce traceplots for specified parameters.

```{r message = FALSE}
library(bayesplot)
mcmc_trace(fit, pars = c("Intercept", "Intercept_sigma"))
```

We also provide a function `epidist_diagnostics()` which can be used to obtain common diagnostics used to assess the quality of a fitted model.

```{r message = FALSE}
epidist_diagnostics(fit)
```

## I'd like to run a simulation study

We recommend use of the [`purrr`](https://purrr.tidyverse.org/) package for running many `epidist` models, for example as a part of a simulation study.
We particularly highlight two functions which might be useful:

1. `purrr::map()` (and other similar functions) for iterating over a list of inputs.
2. `purrr::safely()` which ensures that the function called "always succeeds". In other words, if there is an error it will be captured and output, rather than ending computation (and potentially disrupting a call to `purrr::map()`).

For an example use of these functions, have a look at the [`epidist-paper`](https://github.com/parksw3/epidist-paper) repository containing the code for @park2024estimating.
(Note that in that codebase, we use `map` as a part of a [`targets`](https://books.ropensci.org/targets/) pipeline.)

## How are the default priors for `epidist` chosen?

[`brms`](http://paulbuerkner.com/brms/) provides default priors for all parameters.
However, some of those priors do not make sense in the context of our application.
Instead, we used [prior predictive checking](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html) to set `epidist`-specific default priors which produce epidemiological delay distribution mean and standard deviation parameters in a reasonable range.

For example, for the `brms::lognormal()` latent individual model, we suggest the following prior distributions for the `brms` `mu` and `sigma` intercept parameters:

```{r}
# Note that we export lognormal() as part of epidist hence no need for brms::
family <- lognormal()

epidist_family <- epidist_family(data, family)
epidist_formula <- epidist_formula(
  data,
  family = epidist_family,
  formula = mu ~ 1
)

# NULL here means no replacing priors from the user!
epidist_prior <- epidist_prior(
  data = data,
  family = family,
  formula = epidist_formula,
  prior = NULL
)

epidist_prior
```

(Note that the functions `epidist_family()` and `epidist_prior()` are mostly for internal use!)

Here are the distributions on the delay distribution mean and standard deviation parameters that these prior distributions imply:

```{r message=FALSE, results='hide'}
set.seed(1)
fit_ppc <- epidist(
  data = data,
  formula = mu ~ 1,
  family = lognormal(),
  sample_prior = "only",
  seed = 1,
  backend = "cmdstanr"
)
```

```{r}
pred <- predict_delay_parameters(fit_ppc)

pred |>
  as.data.frame() |>
  pivot_longer(
    cols = c("mu", "sigma", "mean", "sd"),
    names_to = "parameter",
    values_to = "value"
  ) |>
  filter(parameter %in% c("mean", "sd")) |>
  ggplot(aes(x = value, y = after_stat(density))) +
  geom_histogram() +
  facet_wrap(. ~ parameter, scales = "free") +
  labs(x = "", y = "Density") +
  theme_minimal() +
  scale_x_log10(labels = comma)

quantile(pred$mean, c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99))
quantile(pred$sd, c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99))
```

## How can I assess how sensitive the fitted posterior distribution is to the prior distribution used?

We recommend use of the [`priorsense`](https://github.com/n-kall/priorsense) package [@kallioinen2024detecting] to check how sensitive the posterior distribution is to perturbations of the prior distribution and likelihood using power-scaling analysis:

```{r}
library(priorsense)
powerscale_plot_dens(fit, variable = c("Intercept", "Intercept_sigma")) +
  theme_minimal()
```

# What do the parameters in my model output correspond to?

The `epidist` package uses [`brms`](http://paulbuerkner.com/brms/) to fit models.
This means that the model output will include `brms`-style names for parameters.
Here, we provide a table giving the correspondence between the distributional parameter names used in `brms` and those used in standard R functions for some common likelihood families.

| Family        | `brms` parameter | R parameter |
|---------------|------------------|-------------|
| `lognormal()` | `mu`             | `meanlog`   |
| `lognormal()` | `sigma`          | `sdlog`     |
| `Gamma()`     | `mu`             | `shape / scale` |
| `Gamma()`     | `shape`          | `shape`     |
| `weibull()`   | `mu`             | `scale * gamma(1 + 1 / shape)` |
| `weibull()`   | `shape`          | `shape`     |

Note that all families in `brms` are parameterised with some measure of centrality `mu` as their first parameter.
This parameter does not necessarily correspond to the mean: hence the provision of a function `add_mean_sd()` within `epidist` to add columns containing the natural scale mean and standard deviation to a `data.frame` of draws.

# How can I generate predictions with my fitted `epidist` model?

It is possible to generate predictions manually by working with [samples from the model output](https://epidist.epinowcast.org/articles/faq.html#i-would-like-to-work-with-the-samples-output).
However this is tricky to do, and so where possible we recommend using the [`tidybayes`](http://mjskay.github.io/tidybayes/) package.
In particular, following functions may be useful:

1. `tidybayes::add_epred_draws()` for predictions of the expected value of a delay.
2. `tidybayes::add_linpred_draws()` for predictions of the delay distributional parameter linear predictors.
3. `tidybayes::add_predicted_draws()` for predictions of the observed delay.

To see these functions demonstrated in a vignette, see ["Advanced features with Ebola data"](https://epidist.epinowcast.org/articles/ebola.html).
As a short example, to generate 4000 predictions (equal to the number of draws) of the delay that would be observed with a double censored observation process (in which the primary and secondary censoring windows are both one) then:

```{r}
library(tidybayes)
draws_pmf <- tibble::tibble(
  relative_obs_time = Inf, pwindow = 1, swindow = 1, delay_upr = NA
) |>
  add_predicted_draws(fit)

ggplot(draws_pmf, aes(x = .prediction)) +
  geom_bar(aes(y = after_stat(count / sum(count)))) +
  labs(x = "Delay", y = "PMF") +
  scale_x_continuous(limits = c(0, 30)) +
  theme_minimal()
```

Importantly, this functionality is only available for `epidist` models using `brms` families that have a `log_lik_censor` method implemented internally in `brms`. If you are using another family, consider [submitting a pull request](https://github.com/epinowcast/epidist/pulls) to implement these methods!

## How can I use the `marginaleffects` package with my fitted `epidist` model?

The [`marginaleffects`](https://marginaleffects.com/) package provides tools for computing and visualising marginal effects, contrasts, and predictions from regression models. It works with `epidist` models because they are built on top of `brms`.

For `epidist` models with covariates, you can use `marginaleffects` to:
- Compute average marginal effects
- Make comparisons between different covariate values
- Visualise model predictions across the range of covariates

Here's a simple example using a model that includes location as a covariate:

```{r}
library(marginaleffects)

# Compare outcomes between location categories (0 and 1)
avg_comparisons(
  fit_location,
  variables = list(location = c(0, 1))
)
```

For more details on the available functions and their use, see the [`marginaleffects` documentation](https://marginaleffects.com/).


## How can I use the `cmdstanr` backend?

The `cmdstanr` backend is typically more performant than the default `rstan` backend.
To use the `cmdstanr` backend, we first need to install CmdStan (see the README for more details). We can check we have everything we need as follows:

```{r}
cmdstanr::cmdstan_version()
```

## Bibliography
