---
title: "Frequently asked questions and tips"
output: 
  bookdown::html_document2:
    toc: true
    fig_caption: yes
    code_folding: show
    number_sections: false
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

Here we provide tips for working with the `epidist` package, and answers to "frequently" asked questions.
If you have a question about using the package, please create an [issue](https://github.com/epinowcast/epidist/issues) and we will endeavour to get back to you soon!

The code block below is provided to facilitate reproduction of this script, if required!

```{r message=FALSE, results='hide', class.source='fold-hide'}
library(epidist)

set.seed(1)

meanlog <- 1.8
sdlog <- 0.5
obs_time <- 25
sample_size <- 200

obs_cens_trunc <- simulate_gillespie(seed = 101) |>
  simulate_secondary(
    meanlog = meanlog,
    sdlog = sdlog
  ) |>
  observe_process() |>
  filter_obs_by_obs_time(obs_time = obs_time)

obs_cens_trunc_samp <-
  obs_cens_trunc[sample(seq_len(.N), sample_size, replace = FALSE)]

data <- as_latent_individual(obs_cens_trunc_samp)
fit <- epidist(data, seed = 1)
```

## I would like to work with the samples output

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the `posterior` package for working with samples from MCMC or other sampling algorithms.

```{r message = FALSE}
library(posterior)
draws <- posterior::as_draws_df(
  fit,
  variable = c("Intercept", "Intercept_sigma")
)
head(draws)
```

## How can I assess if sampling has converged?

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the `bayesplot` package for sampling diagnostic plots.

```{r message = FALSE}
library(bayesplot)
bayesplot::mcmc_trace(fit, pars = c("Intercept", "Intercept_sigma"))
```

## I'd like to run a simulation study

We recommend use of the `purrr` package for running many `epidist` models, for example as a part of a simulation study.
We particularly highlight two functions which might be useful:

1. `purrr::map` (and other similar functions) for iterating over a list of inputs.
2. `purrr::safely` which ensures that the function called "always succeeds". In other words, if there is an error it will be captured and output, rather than ending computation (and potentially disrupting a call to `purrr::map`).

For an example use of these functions, have a look at the [`epidist-paper`](https://github.com/parksw3/epidist-paper) repository containing the code for @park2024estimating.
(Note that in that codebase, we use `map` as a part of a [`targets`](https://books.ropensci.org/targets/) pipeline.)

## Bibliography
