---
title: "Frequently asked questions and tips"
output: 
  bookdown::html_document2:
    toc: true
    fig_caption: yes
    code_folding: show
    number_sections: false
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

Here we provide tips for working with the `epidist` package, and answers to "frequently" asked questions.
If you have a question about using the package, please create an [issue](https://github.com/epinowcast/epidist/issues) and we will endeavour to get back to you soon!

The code block below is provided to facilitate reproduction of this script, if required!

```{r message=FALSE, results='hide', class.source='fold-hide'}
library(epidist)
library(dplyr)
library(ggplot2)

set.seed(1)

meanlog <- 1.8
sdlog <- 0.5
obs_time <- 25
sample_size <- 200

obs_cens_trunc <- simulate_gillespie(seed = 101) |>
  simulate_secondary(
    meanlog = meanlog,
    sdlog = sdlog
  ) |>
  observe_process() |>
  filter_obs_by_obs_time(obs_time = obs_time)

obs_cens_trunc_samp <-
  obs_cens_trunc[sample(seq_len(.N), sample_size, replace = FALSE)]

data <- as_latent_individual(obs_cens_trunc_samp)
fit <- epidist(data, formula = brms::bf(mu ~ 1, sigma ~ 1), seed = 1)
```

## I would like to work with the samples output

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the `posterior` package for working with samples from MCMC or other sampling algorithms.

```{r message = FALSE}
library(posterior)
draws <- posterior::as_draws_df(
  fit,
  variable = c("Intercept", "Intercept_sigma")
)
head(draws)
```

## How can I assess if sampling has converged?

The output of a call to `epidist` is compatible with typical Stan workflows.
We recommend use of the `bayesplot` package for sampling diagnostic plots.

```{r message = FALSE}
library(bayesplot)
bayesplot::mcmc_trace(fit, pars = c("Intercept", "Intercept_sigma"))
```

## I'd like to run a simulation study

We recommend use of the `purrr` package for running many `epidist` models, for example as a part of a simulation study.
We particularly highlight two functions which might be useful:

1. `purrr::map` (and other similar functions) for iterating over a list of inputs.
2. `purrr::safely` which ensures that the function called "always succeeds". In other words, if there is an error it will be captured and output, rather than ending computation (and potentially disrupting a call to `purrr::map`).

For an example use of these functions, have a look at the [`epidist-paper`](https://github.com/parksw3/epidist-paper) repository containing the code for @park2024estimating.
(Note that in that codebase, we use `map` as a part of a [`targets`](https://books.ropensci.org/targets/) pipeline.)

## How did you choose the default priors for `epidist`?

`brms` provides default priors for all parameters.
However, some of those priors do not make sense in the context of our application.
Instead, we used [prior predictive checking](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html) to set `epidist`-specific default priors which produce epidemiological delay distribution mean and standard deviation parameters in a reasonable range.

For example, for the lognormal latent individual model, we suggest the following prior distributions for the `brms` `mu` and `sigma` intercept parameters:

```{r}
family <- brms::lognormal()

epidist_family <- epidist_family(data, family)
epidist_formula <- epidist_formula(
  data, family = epidist_family, formula = brms::bf(mu ~ 1, sigma ~ 1)
)

# NULL here means no replacing priors from the user!
epidist_prior <- epidist_prior(
  data = data,
  family = family,
  formula = epidist_formula,
  prior = NULL
)

epidist_prior
```

(Note that the functions `epidist_family` and `epidist_prior` are mostly for internal use!)

Here are the distributions on the delay distribution mean and standard deviation parameters that these prior distributions imply:

```{r message=FALSE, results='hide'}
set.seed(1)
fit_ppc <- epidist(
  data = data,
  formula = brms::bf(mu ~ 1, sigma ~ 1),
  family = brms::lognormal(),
  sample_prior = "only",
  seed = 1
)
```

```{r}
pred <- predict_delay_parameters(fit_ppc)

pred |>
  draws_to_long() |>
  filter(parameter %in% c("mean", "sd")) |>
  ggplot(aes(x = value, y = after_stat(density))) +
  geom_histogram() +
  facet_wrap(. ~ parameter, scales = "free") +
  labs(x = "", y = "Density") +
  theme_minimal() +
  scale_x_log10(labels = scales::comma)

quantile(pred$mean, c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99))
quantile(pred$sd, c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99))
```

# What do the parameters in my model output correspond to?

The `epidist` package uses `brms` to fit models.
This means that the model output will include `brms`-style names for parameters.
Here, we provide a table giving the correspondence between the distributional parameter names used in `brms` and those used in standard R functions for some common likelihood families.

| Family        | `brms` parameter | R parameter |
|---------------|------------------|-------------|
| `lognormal()` | `mu`             | `meanlog`   |
| `lognormal()` | `sigma`          | `sdlog`     |
| `Gamma()`     | `mu`             | `shape / scale` |
| `Gamma()`     | `shape`          | `shape`     |

Note that all families in `brms` are parameterised with some measure of centrality `mu` as their first parameter.
This parameter does not necessarily correspond to the mean: hence the provision of a function `add_mean_sd` within `epidist` to add columns containing the natural scale mean and standard deviation to a `data.frame` of draws.

## Bibliography
