---
title: "Left truncation with delay_min"
description: |
  "How to use the delay_min argument to account for delays that are only
  observed above a minimum value."
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
    number_sections: true
pkgdown:
  as_is: true
link-citations: true
vignette: >
  %\VignetteIndexEntry{Left truncation with delay_min}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = file.path("figures", "left-truncation-"),
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

Some delays are only ever recorded above a minimum value.
Generation intervals are often defined to exclude same day transmission.
Case definitions sometimes require a minimum time between two events before a pair is recorded.
In both settings the observed delays are a left truncated sample from the delay distribution of interest.

The `delay_min` argument of `as_epidist_marginal_model()` handles this.
It sets the `L` argument of the [`primarycensored`](https://primarycensored.epinowcast.org/) likelihood, which renormalises the delay distribution over the interval that could have been observed.
Ignoring left truncation biases estimates because the model has to explain the missing short delays by shifting the whole distribution.

`delay_min` truncates on the same scale as the response, that is the observed delay between the primary window lower bound and the secondary event.
Data should therefore be excluded on the observed delay rather than on the underlying continuous delay.

# Setup

```{r load-packages}
library(epidist)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidybayes)
```

# Simulate left truncated data

We simulate an outbreak and a lognormal delay from each primary event.
The parameters are chosen so that a large share of the distribution falls below the truncation point.

```{r simulate}
set.seed(101)

true_meanlog <- 1.0
true_sdlog <- 0.8
delay_min <- 2

obs <- simulate_gillespie(r = 0.2, seed = 101) |>
  simulate_secondary(
    dist = rlnorm,
    meanlog = true_meanlog,
    sdlog = true_sdlog
  )
```

Roughly this share of the delay distribution lies below the truncation point.

```{r truncated-share}
plnorm(delay_min, true_meanlog, true_sdlog)
```

We add daily censoring windows and then drop the pairs whose observed delay is below `delay_min`.

```{r truncate}
obs_trunc <- obs |>
  mutate(
    ptime_lwr = floor(ptime),
    ptime_upr = ptime_lwr + 1,
    stime_lwr = floor(stime),
    stime_upr = stime_lwr + 1
  ) |>
  filter(stime_lwr - ptime_lwr >= delay_min) |>
  mutate(obs_time = max(stime_upr)) |>
  slice_sample(n = 500)
```

The observed delays are truncated below `delay_min`.

```{r hist, fig.cap="Observed delays after left truncation. The dashed line marks the minimum delay."}
ggplot(obs_trunc, aes(x = stime_lwr - ptime_lwr)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 1, fill = "#56B4E9", alpha = 0.7
  ) +
  geom_vline(
    xintercept = delay_min, linetype = "dashed", linewidth = 0.8
  ) +
  labs(x = "Observed delay (days)", y = "Density") +
  theme_minimal()
```

# Fit models with and without the adjustment

```{r prepare}
linelist <- as_epidist_linelist_data(
  obs_trunc$ptime_lwr,
  ptime_upr = obs_trunc$ptime_upr,
  stime_lwr = obs_trunc$stime_lwr,
  stime_upr = obs_trunc$stime_upr,
  obs_time = obs_trunc$obs_time
)

marginal_no_trunc <- as_epidist_marginal_model(linelist)
marginal_trunc <- as_epidist_marginal_model(linelist, delay_min = delay_min)
```

`delay_min` is stored as a column and passed to the likelihood.

```{r show-data}
head(marginal_trunc[, c("delay_lwr", "delay_upr", "delay_min", "n")])
```

```{r fit}
fit_no_trunc <- epidist(
  marginal_no_trunc,
  chains = 2, cores = 2, refresh = ifelse(interactive(), 250, 0)
)

fit_trunc <- epidist(
  marginal_trunc,
  chains = 2, cores = 2, refresh = ifelse(interactive(), 250, 0)
)
```

# Compare parameter estimates

`predict_delay_parameters()` returns posterior draws of the distributional parameters.
For the lognormal family `mu` is the log scale mean and `sigma` is the log scale standard deviation.
We take the first row of the transformed data because there are no covariates here.

```{r compare-params}
param_draws <- list(
  "No adjustment" = fit_no_trunc,
  "With delay_min" = fit_trunc
) |>
  lapply(predict_delay_parameters) |>
  bind_rows(.id = "model") |>
  filter(index == 1) |>
  pivot_longer(
    cols = c("mu", "sigma"),
    names_to = "parameter",
    values_to = "value"
  )

true_params <- data.frame(
  parameter = c("mu", "sigma"),
  value = c(true_meanlog, true_sdlog),
  stringsAsFactors = FALSE
)
```

```{r params-plot, fig.cap="Posterior draws of the lognormal parameters. Dashed lines are the simulation values. The unadjusted model is biased away from them."}
ggplot(param_draws, aes(x = value, fill = model)) +
  geom_density(alpha = 0.6, colour = NA) +
  geom_vline(
    data = true_params, aes(xintercept = value), linetype = "dashed"
  ) +
  facet_wrap(~parameter, scales = "free") +
  scale_fill_manual(values = c(
    "No adjustment" = "#56B4E9",
    "With delay_min" = "#E69F00"
  )) +
  labs(x = "Estimate", y = "Density", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The unadjusted model shifts the distribution to the right and understates its spread.
The adjusted model recovers the simulation values.

# Compare the estimated delay distributions

We predict from both models with `delay_min = 0` so that the predictions describe the underlying delay distribution rather than the truncated one.

```{r predict}
pred_data <- data.frame(
  relative_obs_time = Inf, pwindow = 0, swindow = 0,
  delay_upr = NA, delay_min = 0
)

draws_pred <- bind_rows(
  "No adjustment" = add_predicted_draws(
    pred_data, fit_no_trunc, ndraws = 1000
  ),
  "With delay_min" = add_predicted_draws(
    pred_data, fit_trunc, ndraws = 1000
  ),
  .id = "model"
)
```

```{r pdf-plot, fig.cap="Predicted delay distributions against the true lognormal density (black line)."}
ggplot(draws_pred, aes(x = .prediction)) +
  geom_density(aes(col = model), linewidth = 0.8) +
  geom_function(
    fun = dlnorm,
    args = list(meanlog = true_meanlog, sdlog = true_sdlog),
    linewidth = 1
  ) +
  scale_colour_manual(values = c(
    "No adjustment" = "#56B4E9",
    "With delay_min" = "#E69F00"
  )) +
  coord_cartesian(xlim = c(0, 20)) +
  labs(x = "Delay (days)", y = "Density", col = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Other ways to supply delay_min

Aggregate data works the same way.

```{r aggregate}
agg_data <- as_epidist_aggregate_data(linelist)

marginal_agg <- as_epidist_marginal_model(agg_data, delay_min = delay_min)

head(marginal_agg[, c("delay_lwr", "delay_upr", "delay_min", "n")])
```

When the truncation point varies between observations, pass a column name instead of a number.

```{r per-obs}
linelist_varying <- linelist
linelist_varying$my_min <- rep_len(c(0, 2), nrow(linelist_varying))

marginal_varying <- as_epidist_marginal_model(
  linelist_varying,
  delay_min = "my_min"
)

table(marginal_varying$delay_min)
```

A `delay_min` column already present in the data is used automatically, so `as_epidist_marginal_model()` can be called without any extra argument.

# Summary

Set `delay_min` whenever delays below a threshold could not have been observed.
The default of 0 leaves the likelihood unchanged, so existing models are unaffected.
Note that `delay_min` must not exceed the smallest observed delay, otherwise the data are impossible under the model and `as_epidist_marginal_model()` will error.
