---
title: "Left truncation with delay_min"
description: "Using delay_min to exclude delays below a threshold"
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
    number_sections: true
pkgdown:
  as_is: true
link-citations: true
vignette: >
  %\VignetteIndexEntry{Left truncation with delay_min}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = file.path("figures", "left-truncation-"),
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

Some delay distributions have a natural lower bound above zero.
For example, generation intervals (time between successive infections) are often defined to exclude day zero, as same-day transmission may not be meaningful for a given pathogen.
The `delay_min` parameter in `as_epidist_marginal_model()` supports this by left-truncating the delay distribution at a specified minimum value.
This is passed as the `L` parameter to the [`primarycensored`](https://primarycensored.epinowcast.org/) likelihood.

In this vignette, we demonstrate how to use `delay_min` by simulating data with a known left truncation point and fitting models with and without the truncation adjustment.

# Setup

```{r load-packages}
library(epidist)
library(ggplot2)
library(dplyr)
library(tidybayes)
```

# Simulate data with left truncation

We use the package's simulation tools to generate outbreak data, then remove all observations with delays below a threshold to mimic left truncation.

```{r simulate}
true_meanlog <- 1.5
true_sdlog <- 0.6
delay_min <- 1

outbreak <- simulate_gillespie(r = 0.2, seed = 101)
obs <- simulate_secondary(
  outbreak,
  dist = rlnorm,
  meanlog = true_meanlog,
  sdlog = true_sdlog
)

# Apply left truncation: remove delays below threshold
obs_trunc <- obs |>
  filter(delay >= delay_min) |>
  mutate(
    ptime_lwr = floor(ptime),
    ptime_upr = ptime_lwr + 1,
    stime_lwr = floor(stime),
    stime_upr = stime_lwr + 1,
    obs_time = max(stime_upr)
  ) |>
  filter(stime_upr <= obs_time) |>
  slice_sample(n = 200)
```

The observed delay distribution is visibly truncated at `delay_min = `r delay_min``:

```{r hist, fig.cap="Observed delays are truncated below the minimum delay threshold (dashed line)."}
ggplot(obs_trunc, aes(x = stime_lwr - ptime_lwr)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 1, fill = "#56B4E9", alpha = 0.7
  ) +
  geom_vline(
    xintercept = delay_min, linetype = "dashed", linewidth = 0.8
  ) +
  labs(x = "Observed delay (days)", y = "Density") +
  theme_minimal()
```

# Prepare data and fit models

We convert the simulated data into an `epidist` linelist and prepare marginal models with and without the `delay_min` adjustment.

```{r prepare-and-fit}
linelist <- as_epidist_linelist_data(
  obs_trunc$ptime_lwr,
  ptime_upr = obs_trunc$ptime_upr,
  stime_lwr = obs_trunc$stime_lwr,
  stime_upr = obs_trunc$stime_upr,
  obs_time = obs_trunc$obs_time
)

# Without left truncation adjustment
fit_no_trunc <- epidist(
  as_epidist_marginal_model(linelist),
  chains = 4, cores = 2, refresh = ifelse(interactive(), 250, 0)
)

# With left truncation adjustment
fit_trunc <- epidist(
  as_epidist_marginal_model(linelist, delay_min = delay_min),
  chains = 4, cores = 2, refresh = ifelse(interactive(), 250, 0)
)
```

# Compare parameter estimates

```{r compare-params}
true_params <- data.frame(
  parameter = c("meanlog", "sdlog"),
  true_value = c(true_meanlog, true_sdlog),
  stringsAsFactors = FALSE
)

param_summary <- bind_rows(
  mutate(
    predict_delay_parameters(fit_no_trunc),
    model = "No truncation adjustment"
  ),
  mutate(
    predict_delay_parameters(fit_trunc),
    model = "With delay_min"
  )
) |>
  filter(parameter %in% c("meanlog", "sdlog"))
```

```{r params-plot, fig.cap="Posterior estimates of the lognormal parameters. The model accounting for left truncation (orange) recovers the true values (dashed lines) while the unadjusted model (blue) is biased."}
ggplot(param_summary, aes(x = mean, y = model, col = model)) +
  geom_point(size = 3) +
  geom_linerange(aes(xmin = q5, xmax = q95)) +
  geom_vline(
    data = true_params,
    aes(xintercept = true_value),
    linetype = "dashed"
  ) +
  facet_wrap(~parameter, scales = "free_x") +
  scale_colour_manual(values = c(
    "No truncation adjustment" = "#56B4E9",
    "With delay_min" = "#E69F00"
  )) +
  labs(x = "Estimate", y = "", col = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Compare fitted distributions

```{r predict}
pred_data <- data.frame(
  relative_obs_time = Inf, pwindow = 0, swindow = 0,
  delay_upr = NA, delay_min = 0
)

pred_data_trunc <- data.frame(
  relative_obs_time = Inf, pwindow = 0, swindow = 0,
  delay_upr = NA, delay_min = delay_min
)

draws_no_trunc <- add_predicted_draws(
  pred_data, fit_no_trunc, ndraws = 1000
)

draws_trunc <- add_predicted_draws(
  pred_data_trunc, fit_trunc, ndraws = 1000
)
```

```{r pdf-plot, fig.cap="Predicted delay distributions compared with the true lognormal density (black line). The left-truncated model correctly recovers the distribution shape above the truncation point."}
draws_combined <- bind_rows(
  mutate(draws_no_trunc, model = "No truncation adjustment"),
  mutate(draws_trunc, model = "With delay_min")
)

ggplot(draws_combined, aes(x = .prediction)) +
  geom_density(aes(col = model), linewidth = 0.8) +
  geom_function(
    fun = dlnorm,
    args = list(meanlog = true_meanlog, sdlog = true_sdlog),
    linewidth = 1, linetype = "solid"
  ) +
  scale_colour_manual(values = c(
    "No truncation adjustment" = "#56B4E9",
    "With delay_min" = "#E69F00"
  )) +
  coord_cartesian(xlim = c(0, 30)) +
  labs(x = "Delay (days)", y = "Density", col = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Using delay_min with aggregate data

Left truncation also works with aggregate data.

```{r aggregate}
agg_data <- as_epidist_aggregate_data(linelist)

marginal_agg <- as_epidist_marginal_model(
  agg_data, delay_min = delay_min
)

head(marginal_agg[, c("delay_lwr", "delay_upr", "delay_min", "n")])
```

# Per-observation delay_min

When the truncation point varies across observations, provide `delay_min` as a column name.

```{r per-obs}
linelist_varying <- linelist
linelist_varying$my_min <- sample(
  c(0, 1), nrow(linelist_varying), replace = TRUE
)

marginal_varying <- as_epidist_marginal_model(
  linelist_varying, delay_min = "my_min"
)

table(marginal_varying$delay_min)
```

# Summary

The `delay_min` parameter provides a simple way to account for left truncation when estimating delay distributions.
When delays below a threshold are excluded from the data (as is common for generation intervals), ignoring this truncation biases parameter estimates.
Setting `delay_min` corrects for this by adjusting the likelihood via the [`primarycensored`](https://primarycensored.epinowcast.org/) package.
