---
title: "Guide to the statistical models implemented in epidist"
output: 
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
    number_sections: true
pkgdown:
  as_is: true
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
# exclude compile warnings from cmdstanr
knitr::opts_chunk$set(
  fig.path = "figures/epidist-",
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

The `epidist` package enables users to estimate delay distributions while accounting for common reporting biases.
This vignette describes the models implemented in `epidist` and how they work.

The packages used in this vignette are:

```{r load-requirements}
library(epidist)
library(data.table)
library(purrr)
library(ggplot2)
```

# Background {#maths}

This section introduces the background [@park2024estimating] required to understand the models implemented in `epidist`.

Let $p$ be time of primary event, and $s$ be time of secondary event.
We assume that these events occur at rates, or equivalently incidences, $\mathcal{P}(p)$ and $\mathcal{S}(s)$, respectively.
Let $\tau$ be a delay time such that $\tau = s - p$.
Let $f_p(\tau)$ be probability of delay $\tau$ after observing primary event at time $p$.
Similarly, let $b_s(\tau)$ be probability of delay $\tau$ after observing secondary event at time $s$.
Finally, let $\mathcal{T}(p, s)$ be density of individuals with primary event at time $p$ and secondary event at time $s$.
Then,
$$
\mathcal{T}(p, s) = \mathcal{P}(p) f_p(s - p),
$$
and
$$
\mathcal{T}(p, s) = \mathcal{S}(s) b_s(s - p).
$$
So, by equivalence of the forwards and backwards perspectives then
$$
\mathcal{P}(p) f_p(s - p) = \mathcal{S}(s) b_s(s - p). (\#eq:forwards-backwards)
$$
Rearranging Equation \@ref(eq:forwards-backwards) gives
$$
b_s(\tau) = \frac{\mathcal{P}(s - \tau) f_{s - \tau}(\tau)}{\mathcal{S}(s)}. (\#eq:forwards-backwards2)
$$
The denominator of Equation \@ref(eq:forwards-backwards2), which corresponds to the incidence of secondary events, may be expressed as the integral over all possible delays
$$
\mathcal{S}(s) = \int_{-\infty}^\infty \mathcal{P}(s - \tau) f_{s - \tau}(\tau) \text{d} \tau,
$$
such that
$$
b_s(\tau) = \frac{\mathcal{P}(s - \tau) f_{s - \tau}(\tau)}{\int_{-\infty}^\infty \mathcal{P}(s - \tau) f_{s - \tau}(\tau) \text{d} \tau}.
$$

## Right truncation

Assume the data are right truncated and we don't observe secondary events past time $T$.
Let $P$ and $S$ be random variables.
Let $F_p$ be the forward cumulative distribution.
Then
\begin{align}
\mathbb{P}(S = P + \tau \, | \, P = p, S < T) &= \frac{\mathbb{P}(S = P + \tau, P = p, S < T)}{\mathbb{P}(P = p, S < T)} \\
&= \frac{\mathbb{P}(S = P + \tau < T \, | \, P = p)\mathbb{P}(P = p)}{\mathbb{P}(S < T \, | \, P = p)\mathbb{P}(P = p)} \\
&= \frac{\mathbb{P}(S = P + \tau < T \, | \, P = p)}{\mathbb{P}(S < T \, | \, P = p)} \\
&= \frac{f_p(\tau)}{\int_0^{T - p} f_p(x) \text{d}x} = \frac{f_p(\tau)}{F_p(T - p)}, \quad p + \tau < T. (\#eq:right-truncation)
\end{align}

Examining Equation \@ref(eq:right-truncation) illustrates that (right) truncation renormalises the density by the values which are possible.
For example, if the distribution $x \sim \text{Unif}(0, 1)$ were right truncated by $T = 0.5$ then $x \sim \text{Unif}(0, 0.5)$.

## Interval censoring

To-do: add section about interval censoring.

# The latent individual model

Let $\mathbf{Y}$ be the data vector and $\boldsymbol{\theta}$ be parameters.
Then
$$ 
\mathcal{L}(\mathbf{Y} \, | \, \boldsymbol{\theta}) = \prod_i \mathbb{P}(S_{L, i} < S < S_{R, i} \, | \, P_{L, i} < P < P_{R, i}, S < T).
$$

<!-- Shouldn't this be a joint likelihood? So it includes the prior? I think I took this from the paper, and the paper might have an issue -->

Using latent variables, together with Equation \@ref(eq:right-truncation), then
\begin{align}
x_i &\sim g_P(x_i \, | \, p_{L, i}, p_{R, i}) \\
y_i &\sim \text{Unif}(s_{L, i}, s_{R, i}) \\
\mathcal{L}(\mathbf{Y} \, | \, \boldsymbol{\theta}) &= \prod_i \left[ \frac{f_x(y_i - x_i)}{\int_{P_{L, i}}^{P_{R, i}} g_P(z \, | \, p_{L, i}, p_{R, i}) F_x(T - z) \text{d}z} \right].
\end{align}

Modelling $g_P$ $\iff$ modelling incidence in primary events.
This is a tough thing to do.

This is the approach of the `latent_individual` model [@ward2022transmission] is to:

1. Assume primary event incidence constant
\begin{align}
x_i &\sim \text{Unif}(p_{L, i}, p_{R, i}) \\
y_i &\sim \text{Unif}(s_{L, i}, s_{R, i}).
\end{align}

2. Assume that censoring interval is narrow such that
$$
\int_{P_{L, i}}^{P_{R, i}} g_P(z \, | \, p_{L, i}, p_{R, i}) F_x(T - z) \text{d}z \approx F_x(T - x_i).
$$

Then
$$
\mathcal{L}(\mathbf{Y}_i \, | \, \boldsymbol{\theta}) = \frac{f_x(y_i - x_i)}{F_x(T - x_i)}.
$$


## Bibliography {-}
