---
title: "Primary event distributions"
description: "Choosing how the primary event is distributed within its censoring window"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Primary event distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# The problem

When an event is only observed within a window, say a day or a week, the model needs an assumption about where in that window the event actually happened.
`epidist` calls this the primary event distribution.

The default assumes the primary event is equally likely at any point in its window.
That is reasonable when incidence is flat across the window.
It is wrong when incidence is growing or shrinking quickly, because then the event is more likely to have happened towards one end of the window.
Ignoring that biases the estimated delay distribution [@charniga2024best].

# Setting the distribution

```{r}
library(epidist)
library(dplyr)

obs <- sierra_leone_ebola_data |>
  slice_sample(n = 200) |>
  as_epidist_linelist_data(
    pdate_lwr = "date_of_symptom_onset",
    sdate_lwr = "date_of_sample_tested"
  )
```

The default is a uniform primary event.

```{r}
uniform <- as_epidist_marginal_model(obs)
```

To tilt the primary event towards the end of its window, use `primary = "expgrowth"` and give the exponential growth rate.

```{r}
growing <- as_epidist_marginal_model(
  obs,
  primary = "expgrowth",
  growth_rate = 0.2
)
```

A negative rate tilts it towards the start of the window, which is what you want during a declining epidemic.

```{r}
declining <- as_epidist_marginal_model(
  obs,
  primary = "expgrowth",
  growth_rate = -0.2
)
```

The growth rate is treated as known rather than estimated.
In practice you would take it from a separate estimate of epidemic growth over the period the data cover.

# Does it matter?

The effect is largest when the censoring windows are wide relative to the delay, and when growth is fast.
With daily windows and a delay of several days it is usually small.

```{r fit, eval = FALSE}
fit_uniform <- epidist(uniform, chains = 2, cores = 2, refresh = 0)
fit_growing <- epidist(growing, chains = 2, cores = 2, refresh = 0)

summary(fit_uniform)
summary(fit_growing)
```

If you are unsure of the growth rate, fit under a few plausible values and check whether your conclusions change.

# Limitations

The growth rate is fixed, not estimated, so its uncertainty does not propagate into the delay estimates.

Only the marginal model supports a non-uniform primary event.
The latent model currently assumes a uniform primary event.

# References
