---
title: "Primary event distributions"
description: "Choosing how the primary event is distributed within its censoring window"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Primary event distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# The problem

When an event is only observed within a window, say a day or a week, the model needs an assumption about where in that window the event actually happened.
`epidist` calls this the primary event distribution.

The default assumes the primary event is equally likely at any point in its window.
That is reasonable when incidence is flat across the window.
It is wrong when incidence is growing or shrinking quickly, because then the event is more likely to have happened towards one end of the window.
Ignoring that biases the estimated delay distribution [@charniga2024best].

# Setting the distribution

```{r}
library(epidist)
library(dplyr)

obs <- sierra_leone_ebola_data |>
  slice_sample(n = 200) |>
  as_epidist_linelist_data(
    pdate_lwr = "date_of_symptom_onset",
    sdate_lwr = "date_of_sample_tested"
  )
```

The default is a uniform primary event.

```{r}
uniform <- as_epidist_marginal_model(obs)
```

Both the marginal and the latent model take the same arguments.

```{r}
uniform_latent <- as_epidist_latent_model(obs)
```

To tilt the primary event towards the end of its window, use `primary = "expgrowth"` and give the exponential growth rate.

```{r}
growing <- as_epidist_marginal_model(
  obs,
  primary = "expgrowth",
  growth_rate = 0.2
)
```

A negative rate tilts it towards the start of the window, which is what you want during a declining epidemic.

```{r}
declining <- as_epidist_marginal_model(
  obs,
  primary = "expgrowth",
  growth_rate = -0.2
)
```

The growth rate is treated as known rather than estimated.
In practice you would take it from a separate estimate of epidemic growth over the period the data cover.

# Does it matter?

The effect is largest when the censoring windows are wide relative to the delay, and when growth is fast.
With daily windows and a delay of several days it is usually small.

```{r fit, eval = FALSE}
fit_uniform <- epidist(uniform, chains = 2, cores = 2, refresh = 0)
fit_growing <- epidist(growing, chains = 2, cores = 2, refresh = 0)

summary(fit_uniform)
summary(fit_growing)
```

The Ebola data above has daily windows and a delay of several days, so the effect there is small.

If you are unsure of the growth rate, fit under a few plausible values and check whether your conclusions change.

# How it is implemented

For the marginal model the primary event distribution is handled by `primarycensored`, which integrates over it directly [@primarycensored].

The latent model instead samples the primary event time.
There the exponential tilt is applied unnormalised.
For an observation whose primary and secondary censoring windows do not overlap the normalising constant runs over the primary window, which is data, so it drops out.
For an overlapping observation the primary event is bounded by the sampled secondary event, so a normalising constant would depend on a parameter, and the unnormalised density is used together with the Jacobian described in `vignette("model")`.
Setting `growth_rate = 0` recovers the uniform case exactly.

# Limitations

The growth rate is fixed, not estimated, so its uncertainty does not propagate into the delay estimates.
If that matters, fit under a range of rates and compare.

# References
